---
title: "STATS 250 Lab 11"
author: "Lab Dream Team"
date: "Week of 11/09/2020"
output:
  html_document:
  toc: true
toc_float: true
toc_collapsed: true
theme: lumen
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require(checkmate))
  install.packages('checkmate')
source("plotT.R")
```

## Learning Objectives

### Statistical Learning Objectives
1. Get experience making confidence intervals for population means
1. Understand hypothesis tests for population means

### R Learning Objectives
1. Interpret R output providing confidence intervals and hypothesis tests for inference on population means.


### Functions covered in this lab
1. `pt()`
1. `qt()`
1. `plotT()` (custom)
1. `qqnorm()`
1. `qqline()`
1. `t.test()`

### Weekly Advice
The function `t.test()` will compute the results from both a hypothesis test and a confidence interval. You should pay close attention to the alternative hypothesis. If you are running a hypothesis test, be sure that the correct direction (`greater`, `less`, `two.sided`) is specified in `t.test()`, otherwise your p-value will be inaccurate. If you are running a confidence interval, be sure that the direction is set to `two.sided`, which can be achieved by leaving the alternative hypothesis as the default. You can double-check your thinking by reading the output - the alternative hypothesis is printed in the 4th line of output.
  
## Lab Demo

### Working with the $t$ distribution

The **t distribution**, like the normal distribution that we've seen before, is bell-shaped. Unlike the normal distribution, though, the t has heavier tails than the normal distribution. This is so that we can capture the increased uncertainty introduced by not knowing $\sigma$, the population mean. 

We can find probabilities related to the t distribution using the `pt()` function (sounds similar to `pnorm()`, huh? That's on purpose!). Just like `pnorm()`, `pt()` will give you the area under the curve to the *left* of the value you provide.

```{r pt1}
pt(q = 1.4, df = 4)
plotT(df = 4, shadeValues = 1.4, direction = "less")
# remember the options for direction are "less", "greater", "inside", and "outside"
```

Also, we can use `qt()` to get percentiles of the t distribution, just like with `qnorm()`:

```{r qt1}
qt(p = .8829, df = 4)
```

Cool. Moving on.

### Checking the Normality Condition

Let's say hi to our penguin friends!

```{r penguins}
penguins <- read.csv("https://raw.githubusercontent.com/STATS250SBI/palmerpenguins/master/inst/extdata/penguins_NArm.csv")
str(penguins)
```

Let's say we want to construct a confidence interval for the population mean flipper length of penguins living in the Palmer Archipelago, or conduct a hypothesis test for that mean. 
The first thing we need to do is make sure that it makes sense to do this!

In order to use our machinery for constructing confidence intervals and performing hypothesis tests for means, we need two conditions to hold. **What are they?**

1. *Condition 1*
2. *Condition 2*

Let's check these conditions!

1. There are 333 penguins in the data set; there are probably more than 3,330 penguins in the Palmer archipelago. We don't know exactly the sampling mechanism, though, so we'll proceed with caution.

2. To check the nearly-normal condition, we need to look at the data! One way to do this is to look at a histogram.

```{r flipperHist}
hist(penguins$flipper_length_mm,
     main = "Histogram of Flipper Length",
     xlab = "Flipper length (mm)",
     col = "peachpuff")
```

*Replace this text with your thoughts about this histogram. Is the condition satisfied?*

Another way we can check the normal distribution is using what's called a **Q-Q plot**. To make a Q-Q plot, we'll use the `qqnorm()` function:

```{r flipperQQ}
qqnorm(penguins$flipper_length_mm)
```

So what does this tell us?

- A Q-Q plot plots the actual data on the y-axis ("sample quantiles"), and what that data would look like if it came from a normal distribution on the x-axis ("theoretical quantiles").
- If the y- and x-axes line up (meaning that the points mostly fall on a straight line with positive slope), we can say the sample comes from a normally-distributed population.

It might help to have R draw a line on the plot for us! Let's remake the Q-Q plot, this time adding a `qqline()` (this is similar to an `abline()` but specifically designed for Q-Q plots).

```{r flipperQQ2}
qqnorm(penguins$flipper_length_mm)
qqline(penguins$flipper_length_mm, col = "blue", lwd = 2)
# the col and lwd arguments to qqline() are totally optional; we just wanted to
# show you that you can use them if you want!
```

Are you surprised by the location of that line? Any time you use `qqnorm()` you should also include `qqline()` -- it's best to not eyeball where that line should be.

Here, we can see that there's pretty significant deviation from the line. This reflects the fact that our data were pretty clearly bimodal in that histogram. So the nearly-normal assumption isn't satisfied.

**BUT:** We've got a big sample size (333). What this tells us is that the nearly-normal assumption is not nearly as important as it would be if we didn't have that sample size. _Because we have a large sample size, we can say that the distribution of sample **means** will be nearly normal, which is what we actually need to proceed._


### The `t.test()` function

Let's say we want to construct a 90% confidence interval for the population mean flipper length of penguins living in the Palmer Archipelago. 

The first thing we should do here is to just look at some summary statistics: what does this data look like?

```{r}
summary(penguins$flipper_length_mm)
```

So the mean flipper length of the penguins in our sample is 201 mm. That will be the center of our confidence interval: remember that a confidence interval looks like $$\text{estimate} \pm \text{(a few)} \times \mathrm{SE}_{\text{estimate}}.$$

We can easily do this using the `t.test()` function. This is similar to `prop_test()`! Woo!

```{r t.test1}
t.test(penguins$flipper_length_mm, conf.level = 0.9)
```

So we estimate with 90% confidence that the population mean flipper length for all penguins in the Palmer Archipelago is between 199.7mm and 202.2mm.

Take a look at that p-value now. That thing is suuuuuper small. `2.2e-16` is as small a number as R can make -- it's zero, for all intents and purposes. But let's look at what hypotheses this is testing. Notice the line that says "`alternative hypothesis: true mean is not equal to 0`". That's telling you R ran a test of 
$$H_0: \mu = 0 \quad \text{vs.} \quad H_a: \mu \neq 0.$$

Does our sample provide any evidence that $\mu = 0$? NO. So our p-value is basically zero! We have extremely strong evidence to say that $\mu \neq 0$. This also makes sense because if the population mean flipper length was zero, would penguins still have flippers? They definitely have flippers, tho. Let's pick a more realistic set of hypotheses. 

Let's test $$H_0: \mu = 199 \quad \text{vs.} \quad H_a: \mu > 199.$$ To do this, we'll provide the `mu` and `alternative` arguments to `t.test()`:

```{r t.test2}
t.test(penguins$flipper_length_mm, mu = 199, alternative = "greater")
```

`mu` specifies the null-hypothesis value to use in the test (think of `mu` as setting $\mu_0$).

Notice that this p-value is much more reasonable and the test statistic (`t`) isn't nearly as large. So our data provide pretty strong evidence that the population mean flipper length is greater than 199 mm. 

**Note:** Just like in Lab 10, you'll be interpreting the results of the confidence interval and the hypothesis test in the same manner. All that has changed from last week to this week is the *parameter* of interest: this week our parameter of interest is a population **mean**. Recall that in order to find a mean, we would need to collect **quantitative** data.

**Another note:** It's pretty clear from our knowledge of this data and the bimodal distribution of flipper length that penguin species is probably related to things here. We ignored this in the demo; in real life, we should not. Bimodal distributions tend to suggest that there's an important subgroup effect happening: we should really investigate this rather than just forge ahead like we did.

## Code Cheat Sheet

The formatting of this will be easier to read in the knitted version of the document.

- **`pt(q, df, lower.tail = TRUE)`**
  - `q` is the x-axis value you want to find an area related to
  - `df` is the degrees of freedom of the $t$ distribution
  - `lower.tail` determines whether `pt()` finds the area to the left or right of `q`. If `lower.tail = TRUE` (the default), it shades to the left. If `lower.tail = FALSE`, it shades to the right.
- **`qt(q, df, lower.tail = TRUE)`**
  - `p` is the probability or area under the curve you want to find an x-axis value for
  - `df` is the degrees of freedom of the $t$ distribution
  - `lower.tail` determines whether `pt()` finds the area to the left or right of `q`. If `lower.tail = TRUE` (the default), it shades to the left. If `lower.tail = FALSE`, it shades to the right.
- **`plotT()`**
  - `df` refers to the degrees of freedom of the distribution to plot. You must provide this value.
  - `shadeValues` is a vector of up to 2 numbers that define the region you want to shade
  - `direction` can be one of `less`, `greater`, `outside`, or `inside`, and controls the direction of shading between `shadeValues`. Must be `less` or `greater` if `shadeValues` has only one element; `outside` or `inside` if two
  - `col.shade` controls the color of the shaded region, defaults to `"cornflowerblue"`
  - `...` lets you specify other graphical parameters to control the appearance of the normal curve (e.g., `lwd`, `lty`, `col`, etc.)
- **`qqnorm(y, ...)`**
  - `y` refers to the variable for which you want to create a Q-Q plot
  - `...` lets you control graphical elements of the plot like `pch`, `col`, etc.
- **`qqline(y, ...)`**
  - `y` refers to the variable for which you created a Q-Q plot
  - `...` lets you control graphical elements of the plot like `pch`, `col`, etc.
  - Function can only be used *after* using `qqnorm()`
- **`t.test(x, alternative = c("two.sided", "less", "greater"), mu = 0, conf.level = 0.95)`**
  - `x` is a vector of data values
  - `alternative` specifies the direction of the alternative hypothesis; must be one of "two.sided", "less", or "greater"
  - `mu` indicates the true value of the mean (under the null hypothesis); defaults to 0
  - `conf.level` is the confidence level to be used in constructing a confidence interval; must be between 0 and 1, defaults to 0.95


## Try It!
Depending on your instruction method (synchronous remote, asynchronous remote), complete the following exercises collaboratively. 

1. **Asynchronous remote:** Join a Piazza lab group for asynchronous collaboration.

**Collaborators:** If applicable, replace this text with the names of anyone you collaborated with on this project.

New York is known as "the city that never sleeps". A random sample of 30 New Yorkers were asked how much sleep they get per night. Do these data provide strong evidence that New Yorkers sleep more or less than 8 hours a night on average?

>**1.** Write the hypotheses:

$H_0$: replace this text with your null hypothesis

$H_a$: replace this text with your alternative hypothesis

> **2.** Check the appropriate conditions to be able to complete the hypothesis test. We've read the data in for you, and called it `nyc`. The variable you're interested in is called `sleepHrs`.

```{r readData2}
nyc <- read.csv("nycSleep.csv", stringsAsFactors = T)
```

```{r tryIt2, error = T}
# Replace this comment with code required for Try It 2. (Remember that this text is a comment, so R ignores it; you can delete it if you want.) If you don't delete it, **start your code on a new line that doesn't start with #**

```

> **3.** Calculate the value of the test statistic, the degrees of freedom, and the p-value. Recap your results below. 

```{r tryIt3, error = T}
# Replace this comment with code required for Try It 3. (Remember that this text is a comment, so R ignores it; you can delete it if you want.) If you don't delete it, **start your code on a new line that doesn't start with #**

```

The value of the test statistic is t = *replace this text with the value of the test statistic*.

The degrees of freedom are *replace this text with the degrees of freedom*.

The p-value is *replace this text with the p-value*.

> **4.** Make a graph to show your test statistic and your p-value.

```{r tryIt4, error = T}
# Replace this comment with code required for Try It 4. (Remember that this text is a comment, so R ignores it; you can delete it if you want.) If you don't delete it, **start your code on a new line that doesn't start with #**

```

> **5.** What is the conclusion of the hypothesis test?
  
Replace this text with your answer to Try It 5.

> **6** If you were to construct a 90% confidence interval that corresponded to this hypothesis test, would you expect 8 hours to be in the interval? Explain your reasoning.

Replace this text with your answer to Try It 6.

## Dive Deeper

> **1.** Use `pt()` to help you find the probability of observing this test statistic or something more extreme, under the null hypothesis. Is this the same as the p-value you found in Try It 3? Why or why not?

```{r diveDeeper1}
# Replace this comment with code required for Dive Deeper 1. (Remember that this text is a comment, so R ignores it; you can delete it if you want.) If you don't delete it, **start your code on a new line that doesn't start with #**

```

Replace this text with your written answer for Dive Deeper 1.

> **2.** Without reruning `t.test()`, we would like to assess whether the data provides strong evidence to support that New Yorkers sleep more or less than 7 hours a night on average. Based on your confidence interval which is found in the output of Try It 3, is there strong evidence to support this claim? Why or why not?

Replace this text with your written answer for Dive Deeper 2.

> **3.** The $t$ distribution is used to approximate the standard normal distribution, N(0, 1). As the degrees of freedom of the distribution increase, the $t$(df) distribution resembles N(0, 1) more closely. At what value of the degrees of freedom do probabilities from the $t$ distribution look "close enough" probabilities from the standard normal distribution? Play around with `pt()` and the `df` argument in the chunk below to explore this. In 1-2 sentences, describe why you chose the `df` value that you did.

```{r DiveDeeper3, error = T}
# The area to the left of 1.645 in the N(0, 1) distribution is about .95
pnorm(1.645, mean = 0, sd = 1)

# Now, try setting df to different values in pt() -- at what value of df do you
# start seeing probabilities that are close enough to the normal probability?
pt(1.645, df = 1)
pt(1.645, df = ____)
pt(1.645, df = ____)
pt(1.645, df = ____)
pt(1.645, df = ____)
```

Replace this text with 1-2 sentences about your choice of df value, what that means about sample size, and whether you find that number surprising.

## Wrap-Up and Submission

At the top of the document, make sure you've changed the `author` field to your name (in quotes!). If you'd like, change the date as well.

**When you've finished the lab, click the Knit button one last time.**


### Submission instructions
#### If you're using RStudio Cloud
1. In the Files pane, check the box next to `lab11report.html`.
2. Click More > Export...
3. Click Download and save the file on your computer in a folder you'll remember and be able to find later. You can just call the file `lab11report.html` or whatever else you'd like (as long as you remember what you called it).

#### If you're using RStudio installed on your computer
1. locate the `lab11report.html` file on your computer. The file will be saved in the location indicated at the top of the files pane.

#### Submission to Canvas

1. Click the "Assignments" panel on the left side of the page. Scroll to find "Lab 11", and open the assignment. Click "Submit Assignment". 
2. Towards the bottom of the page, you'll be able to choose `lab11report.html` from the folder you saved it in from RStudio Cloud or noted if you're using RStudio Desktop. **You will only be able to upload a .html file -- do not upload any other file type.**
  3. Click "Submit Assignment". You're done! 
